Hydraulically-vulnerable trees survive on deep-water access in a tropical forest community

### 2020-10-23 manuscript.rmd
Added high-level M&M setup para

# Materials and Methods
## Study site description
\
We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and data on stable water isotopes on tree species and soil from an independent study of tree water-sourcing depths [@Meinzer:1999]. The 1,560ha island was formed after the river Chagres was dammed in 1914 to create Gatun Lake and complete the Panama Canal. The entire island is forested and classified as a tropical moist forest in the Holdridge Life Zone system. Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014]. Rainfall at BCI is seasonal with a mean annual rainfall of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a four-month long dry season from mid-December through mid-April ($\le$ 100 mm rain month^-1^). Among the tree *species* found in the forest canopy, 28% are identified to be at least partly deciduous during the dry season, but this amounts to a maximum of 6.3% *trees* in the canopy that are deciduous, and the percentage of total forest crown area that is deciduous is estimated at 9.7% [@Condit:2000].

@@

## Mortality analyses and species-specific drought experience
\
We calculated mortality rates for large trees ($\ge$10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$30 m) and average abundance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; *n* = `r length(erd.mrate.sp)`). For each of the seven census intervals ($t$) in the 35-yr record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (% yr^-1^) for two successive censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{time}$, where $N$~cen~$_{_{1}}$ and $D$~cen~$_{_{1}}$ are the total number of trees of a species present in cen~1~ and dead in cen~2~, respectively, in the 50 ha plot and $time$ is the duration based on mean dates of cen~1~ and cen~2~. For each species we calculated mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{t=1}^{6} {Mort_{t}}$.

@@

\   Soil Water Potential dynamics ($\Psi$~soil,z~) by depth z (from 0.1 through 13 m) obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of such extreme years, drought intensity at depth, drought seasonality and duration (Fig. \@ref(fig:swp-droughts)). Soil droughts in census intervals 2000-2005 and 2005-2010 distinctly occurred in the wet season, with one to two extreme soil drought years, depending on depth (Fig. S\@ref(fig:swp-droughts); Fig. \@ref(fig:erd-mort)). Across the simulation period (1990-2015) $\Psi$~soil,z~ remained above -0.5 MPa from depths 1.7 to 13 m (Fig. S\@ref(fig:swp-all-depths)).

@@

## Effective rooting depths (ERD)
\
ERD estimated from the model with K~leaf~ and a multiplicative effect of VPD as explanatory variables (Eq. \@ref(eq:M2)) was strongly related to the measured isotopic signal ($\delta$^2^H), that is, with a high goodness-of-fit (*R*^2^ = 0.65; *p* = 0.027, *n* = `r length(unique(erd.iso$sp))`; Fig. \@ref(fig:erd-iso)). Two of the alternative models tested--with K~leaf~ and LAI, with or without an effect of VPD--had a strong relationship with $\delta$^2^H as well, but one that was based on a smaller set of species (*n* = 4) as fewer species (<20 out of the `r length(erd.sp)` tested) passed the criteria of sufficient species-level goodness-of-fit (Fig. S\@ref(fig:other-erd-mods)). For the rest of the analyses we therefore used ERD estimates from the best-fit model (Eq. \@ref(eq:M-chosen)). Notably, two species, *L. seemannii* and *T. tuberculata*, that @Meinzer:1999 found to have xylem sap $\delta$^2^H values between soil water and groundwater, suggesting that these species sourced most of the water from depths > 1 m, were also identified by our model with ERD > 1 m (2.9 m for both species; Fig. \@ref(fig:erd-iso)).

@@

\   In this forest, modeled effective rooting depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from `r min(df.erd.to.plot$depth)` m to `r max(df.erd.to.plot$depth)` m (Fig. \@ref(fig:sp-erd)).

@@

## Coordination of ERD with other hydraulic traits
\
For the subset of eight species with ERD for which hydraulic traits data were available, species with deeper ERD showed greater maximum stem hydraulic conductance (K~max,stem~, Spearmans' *r* = `r cor.k`, *p* = `r p.k` at $\alpha$ = 0.05; Fig. \@ref(fig:erd-hyd)A), less negative leaf $\Psi$~tlp~ (*r* = `r cor.tlp`, *p* = `r p.tlp`; Fig. \@ref(fig:erd-hyd)B), less negative $\Psi$~88,stem~ and thus greater vulnerability to embolism from cavitation (*r* = `r cor.stem.88`, *p* = `r p.stem.88`; Fig. \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi$~min~ - $\Psi$~88,stem~; *r* = `r cor.hsm`, *p* = `r p.hsm`; Fig. \@ref(fig:erd-hyd)D).

@@

## Effective rooting depths, experienced drought and mortality
\
Species with deeper ERD showed lower mean mortality rates across the 34-yr record (1982-2015; *R*^2^ = 0.11, *p* = 0.075, *n* = `r length(erd.mrate.sp)`; Fig. \@ref(fig:erd-mort)), and this relationship was stronger among evergreen species. Species with shallower ERD had significantly higher mortality rates than deeper ERD species in the census intervals when soil droughts occurrend in the wet season (Fig. S\@ref(fig:erd-mort-census)). Analysis of time spent under significant soil drought showed that species with shallower ERD spent greater proportion of days such that $\Psi$~soil,z=ERD~ < $\Psi$~crit~ (Fig. \@ref(fig:erd-time)). Species with ERD deeper than 1 m did not cross species $\Psi$~crit~ for the entire 25-yr period for which soil water potential $\Psi$~soil,z~ was modeled (1990-2015). Soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`;  median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi$~crit~) of the most sensitive tree species (-0.28 MPa).

Fig. \@ref(fig:erd-mort) here.
Fig. \@ref(fig:erd-time) here.

# Discussion
\
Our analyses revealed that co-occurring large canopy tree species with deeper effective rooting depths (ERD) were associated with higher hydraulic efficiency, but not safety. Species with deeper ERD showed higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species. Nevertheless, deeper ERD species showed lower mortality rates than shallower ERD species over the 35-yr record (1982-2015). Analysis of time spent below ($\Psi$~crit~) in the soil zone matching ERD $\Psi$~soil,z~=~ERD~ showed that this was likely due to decreasing realized risk of hydraulic failure with increasing access to water for species with deeper ERD. We thus demonstrate, for the first time, a link between realized drought experience, hydraulic architecture and long-term mortality outcomes for co-occurring adult tree species.

### Realised drought risk integral to assessing drought-vulnerability
\
Our study brings attention to the need for assessing drought-sensitivity in terms of species realized drought risk. We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins, that is, traits that are commonly identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through a study period of greater recurrence of drought. These results are consistent with the two studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016]. @Johnson:2018fad found that mortality decreased with deep rooting even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi$~88,stem~) and had narrower safety-margins ($\Psi$~min~ - $\Psi$~88,stem~). @Nardini:2015 found that a species sourcing deep water in a Karst system showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi$~88,stem~ and narrower safety margin compared to two species sourcing shallow-water.

\   Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak.2018 in which species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, extreme drought. Their hydrological model revealed that the multi-year drought exhausted the deep soil and even bedrock water availablity. In contrast, our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period, $\Psi$~soil~ below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and droughts, that is above ($\Psi$~crit~) of the most sensitive tree species (-0.28 MPa). The difference in our results are likely attributable to site differences in the precipitation regime. In their dry forest site (Mudumalai National Park, India) precipitation is low: 1095 $\pm$ 296 mm (mean $\pm$ SD) compared to `r round(rain.man.stats$rain)` $\pm$ `r round(rain.man.stats$rain.sd)` mm (`r min(clim$Year)`-`r max(clim$Year)`) at BCI, and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (Penman-Monteith PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (Penman-Monteith PET = `r round(mean(clim$pet.PM))` $\pm$ `r round(sd(clim$pet.PM))` mm;`r min(clim$Year)`-`r max(clim$Year)`; this study, @Guo.2016), and thus at BCI deep soil layers are effectively recharged annually (Fig. S\@ref(fig:swp-all-depths); @Chitra-Tarak.2018). The variation in mortality explained by ERD in our study is low (*R*^2^ = 0.11, *p* = 0.075), and other causes may explain observed mortality [@McDowell.2018].

### Effective rooting depths at the tree community level
\
By estimating ERD for large trees of `r length(erd.sp)` canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species-rich tropical forests. Our chosen ERD model is an advance over the only other inverse model of ERD (@Chitra-Tarak.2018; *n* = 12 species) as we utilize physiologically meaningful representation, fewer parameters and corroboration with tree and soil data on $\delta$^2^H.

@@

Figure \@ref(fig:erd-mort): Mortality rate (mean $\pm$ 1SE) versus Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) for `r length(erd.mrate.sp)` canopy species found on Barro Colorado Island. *R*^2^ and *p*-values are given in panel insets.

@@

```{r erd-mort, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-mort}")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_by rdi.gr_evergreen.jpeg"))
```

@@

```{r erd-mort-census, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-census}Species mortality rate in each census interval during the study period plotted against ERD."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_by rdi.gr.jpeg"))
```

>>>>

# Materials and Methods
\
In the following sections we describe the study site, the models and data used to estimate and evaluate (1) soil water potential dynamics by depth in the whole-soil column and (2) species-specific effective rooting depths. These are followed by analyses to address the proposed hypotheses. Details of all datasets used are described in the Supplementary Information.

## Study site description
\
We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and data on stable water isotopes on tree species and soil from an independent study of tree water-sourcing depths [@Meinzer:1999]. The 1,560ha island was formed after the river Chagres was dammed in 1914 to create Gatun Lake and complete the Panama Canal. The entire island is forested and classified as a tropical moist forest in the Holdridge Life Zone system. Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014]. Rainfall at BCI is seasonal with a mean annual rainfall of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a four-month long dry season from mid-December through mid-April ($\le$ 100 mm rain month^-1^). Among the tree *species* found in the forest canopy, 28% are identified to be at least partly deciduous during the dry season, but this amounts to a maximum of 6.3% *trees* in the canopy that are deciduous, and the percentage of total forest crown area that is deciduous is estimated at 9.7% [@Condit:2000].

@@

## Mortality analyses and species-specific drought experience
\
We calculated mortality rates for large trees ($\ge$10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$30 m) and average abundance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; *n* = `r length(erd.mrate.sp)`). For each of the seven census intervals ($t$) in the 35-yr record (1981-82, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (% yr^-1^) for two successive censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{time}$, where $N$~cen~$_{_{1}}$ and $D$~cen~$_{_{1}}$ are the total number of trees of a species present in cen~1~ and dead in cen~2~, respectively, in the 50 ha plot and $time$ is the duration based on mean dates of cen~1~ and cen~2~. For each species we calculated mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{t=1}^{6} {Mort_{t}}$.

@@

\   Soil Water Potential dynamics ($\Psi$~soil,z~) by depth z (from 0.1 through 13 m) obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of such extreme years, drought intensity at depth, drought seasonality and duration (Fig. \@ref(fig:swp-droughts)). Notably, soil droughts in census intervals 2000-2005 and 2005-2010 distinctly occurred in the wet season, with one to two extreme soil drought years each, depending on depth location (Fig. S\@ref(fig:swp-droughts); Fig. \@ref(fig:erd-mort)). Across the simulation period (1990-2015) $\Psi$~soil,z~ remained above -0.5 MPa from depths 1.7 to 13 m (Fig. S\@ref(fig:swp-all-depths)).

@@

## Effective rooting depths (ERD)
\
ERD estimated from the chosen model with K~leaf~ and a multiplicative effect of VPD as explanatory variables (Eq. \@ref(eq:M2)) was strongly related to the measured isotopic signal ($\delta$^2^H), that is, with a high goodness-of-fit (*R*^2^ = 0.65; *p* = 0.027, *n* = `r length(unique(erd.iso$sp))`; Fig. \@ref(fig:erd-iso)). Two of the alternative models tested--with K~leaf~ and LAI, with or without an effect of VPD--had a strong relationship with $\delta$^2^H as well, but one that was based on a smaller set of species (*n* = 4) as fewer species passed the criteria of sufficient species-level goodness-of-fit; <20 vs. `r length(erd.sp)` species for the chosen model (Fig. S\@ref(fig:other-erd-mods)). For the rest of the analyses we therefore used ERD estimates from the less parimonious, chosen model (Eq. \@ref(eq:M-chosen)).

@@

\   In this forest, modeled effective rooting depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from `r min(df.erd.to.plot$depth)` m to `r max(df.erd.to.plot$depth)` m (Fig. \@ref(fig:sp-erd)). Notably, two species, *L. seemannii* and *T. tuberculata*, that @Meinzer:1999 found to have xylem sap $\delta$^2^H values between soil water and groundwater, suggesting that these species sourced most of the water from depths >1 m, were also identified by our model with ERD >1 m (2.9 m for both species; Fig. \@ref(fig:erd-iso)).

@@

## Coordination of ERD with other hydraulic traits
\
For the subset of eight species with ERD for which hydraulic traits data were available, species with deeper ERD showed greater maximum stem hydraulic conductance (K~max,stem~, Spearmans' *r* = `r cor.k`, *p* = `r p.k`; Fig. \@ref(fig:erd-hyd)A), less negative leaf $\Psi$~tlp~ (Spearmans' *r* = `r cor.tlp`, *p* = `r p.tlp`; Fig. \@ref(fig:erd-hyd)B), less negative $\Psi$~88,stem~ and thus greater vulnerability to embolism from cavitation (Spearmans' *r* = `r cor.stem.88`, *p* = `r p.stem.88`; Fig. \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi$~min~ - $\Psi$~88,stem~; Spearmans' *r* = `r cor.hsm`, *p* = `r p.hsm`; Fig. \@ref(fig:erd-hyd)D).

@@

## Effective rooting depths, experienced drought and mortality
\
Species with deeper ERD had significantly lower mortality rates than deeper ERD species in the census intervals 1985-1990, 2000-2005 and 2005-2010. (*n* = `r nrow(mrate.depth.mean)`, Fig. S\@ref(fig:erd-mort-census)). These relationships were stronger among evergreen species ( *n* = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`; Fig. \@ref(fig:erd-mort-census-evg)). Census intervals 2000-2005 and 2005-2010 were also the periods during which extreme soil droughts distinctly occurred in the wet season, effectively extending the dry season (Fig. \@ref(fig:swp-droughts)). Soil water availability was not modelled for 1985-1990. Analysis of a species' time spent under significant soil drought (proportion of days $\Psi$~soil,z=ERD~ < $\Psi$~crit~) showed that species with shallower ERD spent greater proportion of days below critical hydraulic thresholds (Fig. \@ref(fig:erd-time); S\@ref(fig:erd-time-evg)). Species with ERD deeper than 1 m did not cross species $\Psi$~crit~ for the entire 25-yr period for which soil water potential $\Psi$~soil,z~ was modeled (1990-2015). Soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`;  median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi$~crit~) of the most sensitive tree species (-0.28 MPa).

Fig. \@ref(fig:erd-mort-census-evg) here.
Fig. \@ref(fig:erd-time) here.

# Discussion
\
Our analyses revealed that co-occurring large canopy tree species with deeper effective rooting depths (ERD) were associated with higher hydraulic efficiency, but not safety (Fig. \@ref(fig:erd-hyd)). Species with deeper ERD showed higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species. Nevertheless, deeper ERD species showed lower mortality rates than shallower ERD species over the 35-yr record (1981-2015) (Fig. S\@ref(fig:erd-mort-census), especially the evergreen species (Fig. \@ref(fig:erd-mort-census-evg)). Analysis of time spent below ($\Psi$~crit~) in the soil zone matching ERD $\Psi$~soil,z=ERD~ showed that this was likely due to decreasing realized risk of hydraulic failure with increasing access to water for species with deeper ERD (Fig. \@ref(fig:erd-time); Fig. S\@ref(fig:erd-time-evg)). We thus demonstrate, for the first time, a link between realized drought experience, hydraulic architecture and long-term mortality outcomes for co-occurring adult tree species.

### Realised drought risk integral to assessing drought-vulnerability
\
Our study brings attention to the need for assessing drought-sensitivity in terms of species realized drought risk by accounting for the actual hydrological drought and trees' deep water access, beyond the universal drought indices and trees' above-ground hydraulic vulnerability. We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins, that is, traits that are commonly identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through periods of significant droughts (Figs. \@ref(fig:erd-mort-census-evg),  S\@ref(fig:erd-mort-census), \@ref(fig:swp-droughts)). These results are consistent with a few studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016]. @Johnson:2018fad found that mortality decreased with deep rooting even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi$~88,stem~) and had narrower safety-margins ($\Psi$~min~ - $\Psi$~88,stem~). @Nardini:2015 found that a species sourcing deep water in a Karst system showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi$~88,stem~ and narrower safety margin compared to two species sourcing shallow-water.

\   Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak.2018 in which species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, extreme drought. Their hydrological model revealed that the multi-year drought exhausted the deep soil and even bedrock water availablity. In contrast, our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period, $\Psi$~soil~ below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and droughts, that is above ($\Psi$~crit~) of the most sensitive tree species (-0.28 MPa). The difference in our results are likely attributable to site differences in the precipitation regime. In their dry forest site (Mudumalai National Park, India) precipitation is low: 1095 $\pm$ 296 mm (mean $\pm$ SD) compared to `r round(rain.man.stats$rain)` $\pm$ `r round(rain.man.stats$rain.sd)` mm (`r min(clim$Year)`-`r max(clim$Year)`) at BCI, and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (Penman-Monteith PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (Penman-Monteith PET = `r round(mean(clim$pet.PM))` $\pm$ `r round(sd(clim$pet.PM))` mm;`r min(clim$Year)`-`r max(clim$Year)`; this study, @Guo.2016), and thus at BCI deep soil layers are effectively recharged annually (Fig. S\@ref(fig:swp-all-depths); @Chitra-Tarak.2018). And the kind of droughts that buffered the deep rooted species in our study were recurrent sub-annual droughts that extended the dry season into the wet season. The variation in mortality explained by ERD in our study is upto 40%; other causes may explain observed mortality [@McDowell.2018].

### Effective rooting depths at the tree community level
\
By estimating ERD for large trees of `r length(erd.sp)` canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species-rich tropical forests. Our chosen ERD model is an advance over the only other inverse model of ERD (@Chitra-Tarak.2018; *n* = 12 species) as we employ a physiologically meaningful representation, fewer parameters and corroboration with tree and soil data for $\delta$^2^H.

@@

Figure \@ref(fig:erd-mort-census-evg): Mortality rate (mean $\pm$ 1SE) versus Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) for `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))` evergreen, canopy species found on Barro Colorado Island over seven census intervals (1981-2015). *R*^2^ and *p*-values for linear model fits are given in panel insets. Model fits with SE (gray area) are shown for significant fits at $\alpha = 0.1$.

@@

```{r erd-mort-census-evg, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-mort}")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_by rdi.gr_evergreen.jpeg"))
```

@@

```{r erd-mort-census, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-erd-mort-census}Species mortality rate in each census interval during the study period plotted against ERD. *R*^2^ and *p*-values for linear model fits are given in panel insets. Model fits with SE (gray area) are shown for significant fits at $\alpha = 0.1$."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_by rdi.gr.jpeg"))
```

```{r erd-time-evg, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time-evg}Modelled species Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\\Psi$~soil,z~ was below $\\Psi$~crit~, defined as $\\Psi$~80,leaf~, over 1990-2015, where z is the soil depth matching species ERD--for ", nrow(mrate.depth.mean %>% subset(deciduous == "E")), " canopy species found on Barro Colorado Island. Each circle represents a species. Circles are colored and jittered vertically around the named depths (ticks) for clarity.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr_evergreen.jpeg"))
```
